---
title: "Spillover Power & the Saturation-Design"
subtitle: "AI Chatbot Component ESI en Valores RCT"
author: "Nandita"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose of this document

1.  Shows, with two independent tools, **how much power we have to detect a spillover** under different saturation schemes:
    - the **Baird et al. analytic MDE formulas** (their published R code), and
    - the **`RCT2`** package (Jiang, Imai & Malani 2023)

------------------------------------------------------------------------

# Inputs, notation, and assumptions

```{r tbl-inputs}
library(knitr); library(kableExtra)
inputs <- data.frame(
  Parameter = c("Treatment schools (C)", "Students per school (n)", "Total students",
                "Significance (alpha)", "Power (gamma)",
                "ICC — pregnancy", "ICC — overconfidence",
                "ICC — depression / knowledge", "ICC — test scores",
                "Pregnancy base rate (1 yr / 4 yr)", "Overconfidence mean / SD",
                "Female share (girls-only rows)"),
  Value = c("450", "90", "40,500", "0.05 (two-sided)", "0.80",
            "0.050", "0.065", "0.100", "0.200",
            "0.056 / 0.168", "10.4 / 27.7", "0.50"),
  Source = c("design doc", "design doc (30/grade x 3 grades)", "design doc",
             "convention", "convention",
             "ENHOGAR (conservative)", "2024 pilot (N = 800)",
             "health literature / pilot", "Pruebas Nacionales",
             "WDI / Dupas 2018", "2024 pilot", "design doc"),
  check.names = FALSE)
kable(inputs, caption = "Table 1. Design inputs and data (with sources)") |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

```{r tbl-assumptions}
assum <- data.frame(
  Assumption = c("Spillover channel", "Spillover sizes swept", "Saturation designs",
                 "Debiasing coding", "Estimator (chatbot effect)",
                 "Estimator (spillover)", "Simulation reps"),
  Choice = c("untreated outcome rises by  spillover x (school's treated share)",
             "0, 0.10, 0.20, 0.30 SD  (0 = no spillover)",
             "uniform 50/50; equal thirds 25/50/75; extreme 20/50/80 @ 40/20/40",
             "centered (Deb - 0.5) so the chatbot coefficient is the pooled effect",
             "school fixed effects + cluster-robust SEs",
             "controls for the school share; no school FE (strata FE + cluster SEs)",
             "300 per cell (raise for smoother power)"),
  Type = c("our modelling choice (Baird-style)", "placeholder — replace with pilot",
           "design options compared", "specification choice",
           "standard", "standard", "numerical"),
  check.names = FALSE)
kable(assum, caption = "Table 2. Modelling assumptions (our choices)") |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

```{r tbl-notation}
notation <- data.frame(
  Symbol = c("n", "C", "tau", "sigma", "varN", "varCo", "pi", "f",
             "mu_ind", "p_ind", "j, k", "t_alpha", "t_gamma", "Var_S", "MDSE_S"),
  Meaning = c("students per school", "number of schools",
              "between-school variance", "within-school variance",
              "total individual variance", "clustering penalty",
              "saturation levels (treated share per bin)",
              "share of schools at each saturation",
              "expected treated share in a bin", "expected untreated share in a bin",
              "the two saturation bins compared for the slope",
              "test critical value", "power critical value",
              "sampling variance of the spillover-slope estimate",
              "minimum detectable spillover slope"),
  `Formula / value` = c("90", "450", "= ICC", "= 1 - ICC", "tau + sigma  (= 1)",
              "(n - 1) * tau", "e.g. c(.25,.50,.75)", "e.g. c(1/3,1/3,1/3)",
              "pi * f", "(1 - pi) * f", "the extreme bins (1 and 3)",
              "qt(1 - alpha/2, nC - 3)", "qt(gamma, nC - 3)",
              "[varCo*(1/f_j+1/f_k) + varN*(1/p_j+1/p_k)] / (nC)",
              "(t_alpha + t_gamma) * sqrt(Var_S) / (pi_k - pi_j)"),
  Source = c("design", "design", "definition (= ICC)", "definition",
             "Baird et al.", "Baird et al.", "design choice", "design choice",
             "Baird et al.", "Baird et al.", "our choice",
             "Baird et al.", "Baird et al.", "Baird et al.", "Baird et al."),
  check.names = FALSE)
kable(notation, caption = "Table 3. Notation used in the formulae") |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) 
```

------------------------------------------------------------------------

# Background: the design and why spillovers matter

The chatbot is randomized in two stages:

- **Stage 1 (school level):** schools are assigned to *Regular CSE* vs *CSE + Debiasing*. These are **between-school** comparisons.
- **Stage 2 (individual level):** within each treatment school, students are randomized to chatbot access. This is a **within-school** comparison.
- 

| Estimand | Randomized at | School FE? |
|------------------------|------------------------|------------------------|
| CSE vs control; debiasing vs CSE | school level | **No** — absorbed by FE; analyzed between-school |
| Chatbot direct effect | individual, within school | **Yes** — FE removes between-school noise (precision gain) |
| Spillover (school's treated share) | school level (one value per school) | **No** — absorbed by FE; identified between schools |

------------------------------------------------------------------------

# Baird et al. 2018

The key result: **the power to detect the average (pooled) treatment effect declines exactly as you gain the ability to identify spillover and slope effects.** You cannot maximize both.

Rules they derive:

- If you want **both** the treatment and spillover effects, put **more clusters at the extreme saturations** (e.g. more schools at 20% and 80% than at 40/60%). The needed skew shrinks as the ICC rises.
- If you only want the **slope** (how spillovers change with exposure), you don't need a pure-control group; use extreme symmetric saturations (\~0.1 and 0.9).
- For the **pooled treatment effect**, a simple partial-population design (pure control + one interior saturation) is optimal; spreading saturations *reduces* its power.
- **Bottom line:** if spillovers are likely small and ICC is high, don't saturate.

------------------------------------------------------------------------

# Tool 1 - Baird et al. analytic MDEs (base R)

These are the authors' published functions (base R, no packages)

direct file: <https://pdel.ucsd.edu/_files/R%20Implementation.R>

full software page: <https://pdel.ucsd.edu/about/tools/index.html>

`power_slope()` gives the **minimum detectable spillover slope**; how precisely we can estimate the way spillovers change across saturations.

```{r baird-functions}
# Source: Baird, Bohren, McIntosh & Ozler -- PDEL "R Implementation.R"
#   https://pdel.ucsd.edu/_files/R%20Implementation.R
# Base R only (uses qt()); no packages required. Arguments explained in the table below.

power_slope <- function(n, C, alpha, gamma, tau, sigma, pi, f, j, k) {
  t_alpha <- qt(1 - alpha/2, n*C - 3)
  t_gamma <- qt(gamma, n*C - 3)
  mu_ind <- pi * f                 # E[treated]   per saturation bin
  p_ind  <- (1 - pi) * f           # E[untreated] per saturation bin
  varN  <- tau + sigma             # total individual variance
  varCo <- (n - 1) * tau           # cluster-correlation contribution
  Var_T <- (varCo*(1/f[j] + 1/f[k]) + varN*(1/mu_ind[j] + 1/mu_ind[k])) / (n*C)
  Var_S <- (varCo*(1/f[j] + 1/f[k]) + varN*(1/p_ind[j]  + 1/p_ind[k]))  / (n*C)
  list(MDSE_T = ((t_alpha + t_gamma)/(pi[k] - pi[j])) * sqrt(Var_T),   # treatment slope
       MDSE_S = ((t_alpha + t_gamma)/(pi[k] - pi[j])) * sqrt(Var_S))   # spillover slope
}

# power_pooled(): pooled treatment effect vs a PURE CONTROL (MDE_Tonly) and the
# pooled SPILLOVER LEVEL on untreated students vs a pure control (MDE_S).
# Requires a saturation-0 ("pure control") arm
power_pooled <- function(n, C, alpha, gamma, tau, sigma, pi, f) {
  t_alpha <- qt(1 - alpha/2, n*C - 3); t_gamma <- qt(gamma, n*C - 3)
  mu  <- sum(pi * f); eta <- sum(pi^2 * f)
  psi <- if (pi[1] == 0) f[1] else 0          # share of schools that are pure control
  muS <- 1 - mu - psi
  etaT <- (eta - mu^2)/(1 - psi) - (psi/(1 - psi)^2)*mu^2
  varN <- tau + sigma; varCo <- (n - 1)*tau
  Var  <- 1/(n*C)*(varCo*(1/(psi*(1-psi)) + (1-psi)/(mu ^2)*etaT) + varN*(psi+mu )/(mu *psi))
  VarS <- 1/(n*C)*(varCo*(1/(psi*(1-psi)) + (1-psi)/(muS^2)*etaT) + varN*(psi+muS)/(muS*psi))
  Var_T <- 1/(n*C)*(varCo*(eta-mu^2)/(mu^2*(1-mu)^2) + varN/(mu*(1-mu)))
  list(MDE_T = (t_alpha+t_gamma)*sqrt(Var),       # pooled treatment (needs pure control)
       MDE_S = (t_alpha+t_gamma)*sqrt(VarS),      # pooled spillover LEVEL vs control
       MDE_Tonly = (t_alpha+t_gamma)*sqrt(Var_T)) # treatment vs control
}

# Spillover-slope MDE only (safe when a bin is a pure control: avoids the
# treated-slope term, which divides by zero when a bin has no treated students).
spill_slope_S <- function(n, C, alpha, gamma, tau, sigma, pi, f, j, k) {
  tt <- qt(1 - alpha/2, n*C - 3) + qt(gamma, n*C - 3)
  p_ind <- (1 - pi) * f
  varN <- tau + sigma; varCo <- (n - 1)*tau
  Var_S <- (varCo*(1/f[j] + 1/f[k]) + varN*(1/p_ind[j] + 1/p_ind[k])) / (n*C)
  (tt / (pi[k] - pi[j])) * sqrt(Var_S)
}
```

The spillover effect can differ across saturation levels; its *slope* is (spillover at high saturation − spillover at low saturation) ÷ (high − low share). `power_slope()` returns the smallest such slope we could distinguish from zero: a standard-error term `sqrt(Var_S)`, scaled up by the test-and-power constants `(t_alpha + t_gamma)`, and **divided by the saturation gap** `(pi[k] − pi[j])`.

## Design comparison: equal-thirds vs extreme-weighted

```{r design-comparison}
library(knitr)
library(kableExtra)   # install.packages("kableExtra") if needed

n <- 90; C <- 450
outcomes <- list(c("Pregnancy rate",         0.050),
                 c("Overconfidence score",   0.065),
                 c("Depression / Knowledge", 0.100),
                 c("Test scores",            0.200))

cmp <- do.call(rbind, lapply(outcomes, function(o) {
  icc <- as.numeric(o[2]); tau <- icc; sig <- 1 - icc
  eq <- power_slope(n, C, .05, .80, tau, sig, c(.25,.50,.75), c(1/3,1/3,1/3), 1, 3)$MDSE_S
  ex <- power_slope(n, C, .05, .80, tau, sig, c(.20,.50,.80), c(.40,.20,.40), 1, 3)$MDSE_S
  data.frame(Outcome = o[1],
             ICC = sprintf("%.3f", icc),
             Eq  = sprintf("%.4f", eq),
             Ex  = sprintf("%.4f", ex),
             Gain = sprintf("-%.1f%%", (1 - ex/eq) * 100),
             check.names = FALSE)
}))
colnames(cmp) <- c("Outcome", "ICC", "Equal thirds 25/50/75",
                   "Extreme 20/50/80 (40/20/40)", "MDE reduction")

kable(cmp, align = c("l", "c", "c", "c", "c"),
      caption = paste0("Minimum detectable spillover SLOPE (SD units) by saturation design ",
                       "(80% power, alpha = 0.05, 450 schools x 90 students)")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  add_header_above(c(" " = 2, "MDE of spillover slope (SD)" = 2, " " = 1)) |>
  column_spec(4, bold = TRUE) 
```

The extreme-weighted design (20/50/80, with 40% of schools at each extreme and only 20% at 50%) lowers the detectable spillover by roughly a fifth to a quarter across outcomes. Even so, the detectable spillover is large in absolute terms (about 0.14-0.31 SD) and worst for high-ICC outcomes like test scores, so only sizeable spillovers will be detectable.

## Pure control group addition

In the current design no treatment school is at saturation 0 but we can add one (CSE schools where 0% get the chatbot). Doing so is what unlocks Baird's pure-control formulas and gives two *spillover-clean* comparisons.

| \# | Estimand | What it answers | Tool | Needs pure control? |
|---------------|---------------|---------------|---------------|---------------|
| 1 | **Direct effect (within-FE)** | chatbot vs a *classmate* without it | within-school FE formula | no |
| 2 | **Treatment vs control** | chatbot vs a *pure-control school* (spillover-clean) | `power_pooled()` `MDE_Tonly` | **yes** |
| 3 | **Spillover slope** | how spillover *changes* with saturation | `power_slope()` `MDSE_S` | no (needs ≥2 saturations) |
| 4 | **Spillover level vs control** | untreated-in-treated-school vs pure control | `power_pooled()` `MDE_S` | **yes** |

Estimand 1 is precise but spillover-biased; estimand 2 is the clean version (needs a pure control). Estimand 3 measures the *shape* of spillover across saturations; estimand 4 measures its *level* against a clean baseline.

For estimand 1 the formula is the within-school one, `(t_alpha + t_gamma) * sigma_within * sqrt(1 / (N * mean[s(1-s)]))`; the design enters only through `mean[s(1-s)]`, largest at a 50/50 split.

## Variation across designs

```{r master-menu}
icc  <- 0.20                               # test scores (highest ICC = hardest); edit to compare outcomes
zc   <- qnorm(0.975) + qnorm(0.80)         # 2.80 test+power constant
Ntot <- 90 * 450

# Estimand 1: within-school direct effect (chatbot vs classmate). 0 if no within-school variation.
direct_mde <- function(pi, f) {
  mean_s <- sum(f * pi * (1 - pi))         # avg treatment-assignment variance
  if (mean_s == 0) return(NA)
  zc * sqrt(1 - icc) * sqrt(1 / (Ntot * mean_s))
}

# design menu: name, saturations (pi), school shares (f), spillover-slope bins (j,k), pure control?
menu <- list(
  list("Uniform 50/50",               c(.50),               c(1),            NA,     FALSE),
  list("Partial pop: control + 50%",  c(0,.50),             c(.50,.50),      c(1,2), TRUE),
  list("Equal thirds 25/50/75",       c(.25,.50,.75),       c(1/3,1/3,1/3),  c(1,3), FALSE),
  list("Extreme 20/50/80 (40/20/40)", c(.20,.50,.80),       c(.40,.20,.40),  c(1,3), FALSE),
  list("Symmetric 20/80 (50/50)",     c(.20,.80),           c(.50,.50),      c(1,2), FALSE),
  list("Wide 10/50/90 (40/20/40)",    c(.10,.50,.90),       c(.40,.20,.40),  c(1,3), FALSE),
  list("Baird ladder 0/25/50/75/100", c(0,.25,.50,.75,1),   rep(.2,5),       c(2,4), TRUE),
  list("Control + thirds 0/25/50/75", c(0,.25,.50,.75),     rep(.25,4),      c(2,4), TRUE),
  list("Control + 20/50/80 (0/20/50/80)", c(0,.20,.50,.80), rep(.25,4),      c(2,4), TRUE))

fmt <- function(x) ifelse(is.na(x) | !is.finite(x), "—", sprintf("%.3f", x))
menu_tbl <- do.call(rbind, lapply(menu, function(d) {
  pi <- d[[2]]; f <- d[[3]]; jk <- d[[4]]; pc <- d[[5]]
  slope  <- if (length(pi) > 1) spill_slope_S(90, 450, .05, .80, icc, 1 - icc, pi, f, jk[1], jk[2]) else NA
  pooled <- if (pc) power_pooled(90, 450, .05, .80, icc, 1 - icc, pi, f) else list(MDE_Tonly = NA, MDE_S = NA)
  data.frame(Design = d[[1]], `Pure control?` = ifelse(pc, "Yes", "No"),
             `1. Direct (within-FE)`     = fmt(direct_mde(pi, f)),
             `2. Treatment vs control`   = ifelse(pc, fmt(pooled$MDE_Tonly), "—"),
             `3. Spillover slope`        = fmt(slope),
             `4. Spillover vs control`   = ifelse(pc, fmt(pooled$MDE_S), "—"),
             check.names = FALSE)
}))
kable(menu_tbl, align = c("l","c","c","c","c","c"),
      caption = "Test scores (ICC 0.20): MDE (SD) for each estimand, by design. Lower = more power; — = not identified.") |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) |>
  add_header_above(c(" " = 2, "Treatment effect" = 2, "Spillover" = 2)) |>
  footnote(general = paste0(
    "Estimand 1 (within-FE direct): chatbot vs a classmate without it — precise but spillover-biased. ",
    "Estimand 2 (Baird power_pooled MDE_Tonly): chatbot vs a PURE-CONTROL school — spillover-clean. ",
    "Estimand 3 (Baird power_slope MDSE_S): how spillover changes across saturations. ",
    "Estimand 4 (Baird power_pooled MDE_S): untreated-in-treated-school vs pure control — the spillover LEVEL. ",
    "Estimands 2 and 4 require a saturation-0 (pure-control) arm; hence '—' for designs without one."),
    general_title = "Notes: ")
```

1.  **Direct effect (col 1)** is best under **uniform 50/50** (\~0.025) and gets worse as we add pure-control or extreme schools (they have little/no within-school chatbot variation). But this estimate is *spillover-biased*.
2.  **Treatment vs control (col 2)** is the *spillover-clean* direct effect — available only with a pure control (\~0.085). It's less precise than col 1 (it's a between-school comparison) but immune to contamination. Note it's still a small, detectable effect.
3.  **Spillover slope (col 3)** is best when the two extreme bins are wide *and* hold many schools: wide 10/50/90 is best (\~0.19), then symmetric (\~0.21) and extreme (\~0.24); equal thirds is weaker (\~0.31); and the ladder (\~0.40) and control+thirds (\~0.36) are the worst — spreading schools across many bins leaves only a few in each compared bin, which hurts the slope.
4.  **Spillover level vs control (col 4)** is the best-powered way to detect a spillover at all (\~0.12 with the partial-population design) *because* it leans on the clean pure-control baseline.

Columns 1 and 3 pull in opposite directions (Baird's central result): spreading toward the extremes helps the spillover but costs direct-effect precision. But if the direct effect has some slack (it stays \~0.025–0.085 SD everywhere) then the **spillover is the binding constraint**. Onme straightforward way in which case would be to **add a pure-control arm** (estimand 4, \~0.12 SD), which *also* buys a clean treatment effect (estimand 2).

The same rankings hold for the lower-ICC outcomes (pregnancy, overconfidence); the absolute MDEs just shrink.

------------------------------------------------------------------------

# Tool 2 - `RCT2` cross-check (R package)

Its `Calsamplesize()` returns the **number of clusters** needed to detect a given effect at a target power, per assignment mechanism. The effect-size argument `mu` is *"the largest **direct** effect across treatment assignment mechanisms.*

We run `Calsamplesize` for the saturation designs and compare the clusters it needs against our **450 schools**

```{r rct2-inputs}
rct2_inputs <- data.frame(
  Input = c("Clusters (schools)", "Students per school", "ICC",
            "Direct effect to detect (mu)", "Significance / power",
            "Designs compared", "Excluded", "Method"),
  Value = c("450", "90", "0.10", "0.05 SD", "0.05 / 0.80",
            "the 3-mechanism designs: equal thirds 25/50/75; extreme 20/50/80; wide 10/50/90",
            "uniform 50/50 (1 mechanism) and symmetric 20/80 (2 mechanisms) - Calsamplesize needs 3 arms; also 0%/100% arms (direct effect undefined there)",
            "RCT2 design-based (Jiang-Imai-Malani 2023); data simulated to the assumed ICC"),
  check.names = FALSE)
knitr::kable(rct2_inputs, caption = "RCT2 cross-check: inputs and assumptions") |>
  kableExtra::kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

```{r rct2-designs, eval=FALSE}
library(RCT2); library(kableExtra)
set.seed(2026)
n_per <- 90; n_clust <- 450; icc <- 0.10
mu <- 0.05                          # DIRECT effect to detect (SD); RCT2's 'mu'

# NOTE: RCT2's Calsamplesize is built around a THREE-mechanism design (its worked
# example uses three arms) and its internal indexing errors for 1- or 2-mechanism
# designs. So we compare only the 3-mechanism designs here (uniform = 1 arm and
# symmetric 20/80 = 2 arms are excluded), and it sizes the DIRECT effect, so 0%/100%
# arms don't apply either. For each design the three ASSIGNMENT MECHANISMS are the
# three treated shares a school can be assigned to.
designs_rct2 <- list(
  "Equal thirds 25/50/75" = list(lv = c(.25,.50,.75), w = c(1/3,1/3,1/3)),  # mechanisms: 25% / 50% / 75% treated
  "Extreme 20/50/80"      = list(lv = c(.20,.50,.80), w = c(.40,.20,.40)),  # mechanisms: 20% / 50% / 80% treated
  "Wide 10/50/90"         = list(lv = c(.10,.50,.90), w = c(.40,.20,.40)))  # mechanisms: 10% / 50% / 90% treated

build_dat <- function(lv, w) {
  sat <- sample(lv, n_clust, replace = TRUE, prob = w)   # random mechanism per school (matches the run that worked)
  A   <- rep(sat, each = n_per)                          # A = the school's assignment mechanism (treated share)
  Z   <- rbinom(n_clust * n_per, 1, A)                   # treat each student at the school's rate
  u   <- rep(rnorm(n_clust, 0, sqrt(icc)), each = n_per)
  e   <- rnorm(n_clust * n_per, 0, sqrt(1 - icc))
  Y   <- 0.10 * Z + 0.10 * A * (1 - Z) + u + e
  data.frame(Z = Z, A = A, Y = Y, id = rep(1:n_clust, each = n_per))   # integer id (matches the run that worked)
}

rows <- lapply(names(designs_rct2), function(nm) {
  d   <- designs_rct2[[nm]]
  dat <- build_dat(d$lv, d$w)
  qa  <- d$w / sum(d$w)                                # share of schools per mechanism
  ss  <- Calsamplesize(dat, mu, qa, alpha = 0.05, beta = 0.20)   # class "sample": numeric vector
  cl  <- as.numeric(ss)                                # clusters needed, one per mechanism
  data.frame(Design = nm,
             Mechanisms = paste0(d$lv * 100, "%", collapse = " / "),
             `Clusters needed (per mechanism)` = paste(round(cl), collapse = " / "),
             `Min clusters (best mechanism)`   = round(min(cl)),
             check.names = FALSE)
})
kable(do.call(rbind, rows), align = c("l","c","c","c"),
      caption = "RCT2 (design-based): clusters needed to detect a 0.05 SD DIRECT effect, by saturation design (ICC 0.10)") |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) |>
  footnote(general = paste0(
    "Calsamplesize sizes the DIRECT effect (mu = largest direct effect across mechanisms). ",
    "Compare 'Min clusters' against our 450 schools; the 50% mechanism is the most efficient. ",
    "Numbers are randomization-based and corroborate Baird directionally, not exactly."),
    general_title = "Notes: ")
```

------------------------------------------------------------------------

# Design × spillover grid (simulation)

For each assignment design, and a range of assumed spillover sizes (**including none**), this simulation reports (a) the **bias** of the naive within-school direct-effect estimate, (b) whether the true direct effect can be **recovered**, and (c) the **power to detect the spillover**.

## Assumptions

- **Design held fixed:** 450 treatment schools, 90 surveyed students each.
- **Outcome standardized** so total variance = 1; the **ICC** sets the between-school share of variance (we use 0.10 here).
- **True direct chatbot effect planted at 0.10 SD**
- **Spillover model:** untreated students gain `spillover * (school's treated share)` —{0, 0.10, 0.20, 0.30} SD; **0 = no spillover** (a validity check).
- **Nine designs:** uniform 50/50; partial population (0/50); equal thirds (25/50/75); extreme (20/50/80); symmetric (20/80); wide (10/50/90); control+full (0/50/100); control+thirds (0/25/50/75); and the Baird ladder (0/25/50/75/100).
- **What we measure per design × spillover:** (i) the **bias** of the naive within-school estimate (the attenuation a design would suffer), and (ii) the **power to detect the spillover** by testing whether *untreated* students' outcomes rise with their school's saturation.

```{r grid-setup}
library(fixest)      # fixed-effects regressions + cluster-robust SEs
library(kableExtra)  # formatted table output
set.seed(2026)

S        <- 450                      # treatment schools
n_per    <- 90                       # students surveyed per school
icc      <- 0.10                     # between-school share of variance (ICC)
b2_true  <- 0.10                     # TRUE direct chatbot effect we plant (SD units)
spills   <- c(0, 0.10, 0.20, 0.30)   # spillover sizes to test (0 = none)
REPS     <- 150                      # reps per cell 

# every candidate design as (saturation levels, share of schools at each level)
DESIGN_SPECS <- list(
  "Uniform 50/50"                = list(lv = c(.50),             w = c(1)),
  "Partial pop 0/50"             = list(lv = c(0,.50),           w = c(.50,.50)),
  "Equal thirds 25/50/75"        = list(lv = c(.25,.50,.75),     w = c(1/3,1/3,1/3)),
  "Extreme 20/50/80"             = list(lv = c(.20,.50,.80),     w = c(.40,.20,.40)),
  "Symmetric 20/80"              = list(lv = c(.20,.80),         w = c(.50,.50)),
  "Wide 10/50/90"                = list(lv = c(.10,.50,.90),     w = c(.40,.20,.40)),
  "Control+full 0/50/100"        = list(lv = c(0,.50,1),         w = c(1/3,1/3,1/3)),
  "Control+thirds 0/25/50/75"    = list(lv = c(0,.25,.50,.75),   w = c(.25,.25,.25,.25)),
  "Baird ladder 0/25/50/75/100"  = list(lv = c(0,.25,.50,.75,1), w = rep(.2, 5)),
  "Control + 20/50/80 0/20/50/80" = list(lv = c(0,.20,.50,.80),   w = c(.25,.25,.25,.25)))
designs <- names(DESIGN_SPECS)
```

## Step 1 — assign each school a saturation

`assign_sat()` gives every school its treated share under the chosen design, stratified by CSE arm so the two arms stay balanced.

```{r grid-step1}
assign_sat <- function(design, deb) {
  sp <- DESIGN_SPECS[[design]]; lv <- sp$lv; w <- sp$w
  if (length(lv) == 1) return(rep(lv, S))                   # uniform: every school same share
  sat <- numeric(S)
  for (arm in c(0, 1)) {                                    # stratify within each CSE arm
    idx <- which(deb == arm); m <- length(idx)
    cnt <- floor(w * m); cnt[length(cnt)] <- m - sum(cnt[-length(cnt)])
    lvl <- unlist(mapply(function(c, l) rep(l, c), cnt, lv))
    sat[idx] <- sample(lvl)                                 # shuffle which school gets which
  }
  sat
}
```

## Step 2 — generate virtual study and estimate it two ways

The outcome is built as:

```         
Y = school_effect(ICC) + 0.10*Chatbot + spillover*share*(1 - Chatbot) + noise
```

i.e. a school baseline + the planted direct effect for treated students + a spillover bonus for *untreated* students that grows with their school's treated share + individual noise. We then run both regressions on that one dataset.

```{r grid-step2}
sim_once <- function(design, spill) {
  deb <- sample(rep(c(0, 1), each = S / 2))                 # half schools debiasing
  sat <- assign_sat(design, deb)                            # Step 1
  k   <- round(sat * n_per)                                 # treated count per school
  Tt  <- unlist(lapply(k, function(kk) sample(rep(c(1, 0), c(kk, n_per - kk)))))
  d <- data.frame(school = rep(seq_len(S), each = n_per),
                  s_s = rep(sat, each = n_per), T = Tt)      # s_s = school treated share
  u <- rep(rnorm(S, 0, sqrt(icc)), each = n_per)            # between-school (ICC)
  e <- rnorm(S * n_per, 0, sqrt(1 - icc))                   # within-school
  d$y <- u + b2_true * d$T + spill * d$s_s * (1 - d$T) + e  # plant the outcome

  # (i) NAIVE within-school direct effect (mixed schools only): biased if spillover present
  mid   <- d[d$s_s > 0 & d$s_s < 1, ]
  b2_fe <- coef(feols(y ~ T | school, data = mid, cluster = ~school))[["T"]]

  # (ii) SPILLOVER detection (robust for every design): do UNTREATED students'
  #      outcomes rise with their school's saturation? (cluster-robust by school)
  un <- d[d$T == 0, ]
  reject <- NA_real_
  if (length(unique(un$s_s)) > 1) {
    m <- feols(y ~ s_s, data = un, cluster = ~school)
    reject <- as.numeric(abs(m$coeftable["s_s", "t value"]) > 1.96)
  }
  c(b2_fe = b2_fe, reject = reject)
}
```

## Step 3 — repeat over the whole grid and summarise

```{r grid-step3}
res <- data.frame()
for (dz in designs) for (sp in spills) {
  draws <- replicate(REPS, sim_once(dz, sp))                # 2 x REPS matrix
  res <- rbind(res, data.frame(
    design = dz,
    spill  = sp,
    est    = mean(draws["b2_fe", ]),                        # naive direct estimate (true = 0.10)
    power  = mean(draws["reject", ], na.rm = TRUE)))        # power to detect spillover
}
res
```

## Step 4 — format the results table

```{r grid-step4}
disp <- data.frame(
  `Spillover (delta)`            = sprintf("%.2f", res$spill),
  `Naive direct estimate`        = sprintf("%+.3f", res$est),
  `Spillover detection power`    = ifelse(is.finite(res$power), sprintf("%.0f%%", 100 * res$power), "—"),
  check.names = FALSE)

kt <- kable(disp, align = c("c", "c", "c"),
      caption = sprintf(paste0("Design x spillover grid (live simulation): ",
                               "true direct effect = %.2f SD, ICC = %.2f, %d reps/cell"),
                        b2_true, icc, REPS)) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
ns <- length(spills)
for (i in seq_along(designs)) kt <- pack_rows(kt, designs[i], (i - 1)*ns + 1, i*ns)
kt |> footnote(general = paste0(
    "Naive direct estimate = within-school FE estimate on mixed schools; the TRUE value is 0.10 but it ",
    "is biased by about -0.5*spillover under EVERY design (it collapses to ~0 at spillover 0.20 and flips ",
    "negative at 0.30).",
    "Spillover detection power = share of reps where untreated students' outcomes significantly rise with ",
    "their school's saturation (cluster-robust, alpha = 0.05). Uniform 50/50 has no variation among the ",
    "untreated, so it cannot detect spillover ('—'). The delta = 0 rows are validity checks (~5%). "),
    general_title = "Notes: ")
```

- **Spillover detection power rises with the spillover size**. At a 0.20 spillover, the **spread designs do best** (symmetric 20/80 and wide 10/50/90 ≈ 90%+, control+thirds and partial-pop ≈ 85–90%, extreme ≈ 85%), while **equal thirds (\~68%)** and **control+full 0/50/100 (\~70%)** trail; the latter because its 100%-saturation third has no untreated students to measure spillover on.

------------------------------------------------------------------------

# TBD:

- **Do we want to estimate spillover effects (is that a study objective)?**
  - **If yes:** a saturation design is justified and it should probably include a **pure-control (0%) arm plus several interior levels**.

  - **If no:** don't saturate.

------------------------------------------------------------------------

# Scorecard: both tools at a glance

```{r scorecard}
ref_spill <- 0.20
zc <- qnorm(.975) + qnorm(.80)
score <- do.call(rbind, lapply(designs, function(dz) {
  sp <- DESIGN_SPECS[[dz]]
  meanS <- sum(sp$w * sp$lv * (1 - sp$lv))                 # avg treatment-assignment variance
  me <- if (meanS > 0) zc * sqrt(1 - icc) * sqrt(1/(S*n_per*meanS)) else NA
  pw <- res$power[res$design == dz & res$spill == ref_spill]
  data.frame(Design = dz,
    `Pure control?`            = ifelse(0 %in% sp$lv, "Yes", "No"),
    `Main-effect MDE`          = ifelse(is.na(me), "—", sprintf("%.3f SD", me)),
    `Spillover power (d=0.20)` = ifelse(length(pw) && is.finite(pw), sprintf("%.0f%%", 100*pw), "—"),
    check.names = FALSE)
}))
kable(score, align = c("l","c","c","c"),
      caption = sprintf("Design scorecard (ICC %.2f): main-effect precision (Tool 1) vs spillover detection (grid sim)", icc)) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) |>
  add_header_above(c(" " = 2, "Tool 1 (analytic)" = 1, "Grid (simulation)" = 1)) |>
  footnote(general = paste0(
    "Main-effect MDE: within-school FE formula (Tool 1) - smaller = more precise direct effect. ",
    "Spillover power: grid simulation at a true 0.20 SD spillover - higher = easier to detect. ",
    "The columns trade off: spread/extreme designs detect spillover best at a slightly larger main-effect ",
    "MDE; uniform 50/50 has the smallest main-effect MDE but cannot detect spillover. A pure-control arm ",
    "additionally gives a spillover-CLEAN direct effect (Tool 1 master menu)."),
    general_title = "Notes: ")
```

**Uniform 50/50** has the smallest main-effect MDE (most precise direct effect) but **cannot detect spillover at all**. The **spread designs** (symmetric 20/80 and wide 10/50/90 \~94%, extreme \~86%) and the better **pure-control designs** (partial-pop \~87%, control+thirds \~89%) detect spillover best, for only a small increase in the main-effect MDE.

------------------------------------------------------------------------

# References

- Baird, S., Bohren, J. A., McIntosh, C., & Özler, B. (2018). *Optimal Design of Experiments in the Presence of Interference.* Review of Economics and Statistics, 100(5), 844–860.
- Bohren, Staples, Baird, McIntosh, Özler (2016). *Power Calculation Software for Randomized Saturation Experiments* (PDEL, UCSD) — GUI + Python/R/MATLAB code: <https://pdel.ucsd.edu/about/tools/index.html>
- Jiang, Z., Imai, K., & Malani, A. (2023). *Statistical Inference and Power Analysis for Direct and Spillover Effects in Two-Stage Randomized Experiments.* Biometrics, 79(3), 2370–2383. `RCT2` package: <https://cran.r-project.org/package=RCT2>
- World Bank Development Impact blog (practical summary): <https://blogs.worldbank.org/en/impactevaluations/power-calculation-software-randomized-saturation-experiments> \`\`\`
