---
title: "Power Calculations: AI Chatbot Component — ESI en Valores RCT"
author: "Tomás Buitrago | Development Innovation Lab"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 9,
  fig.height = 5.5,
  fig.align  = "center"
)
```

```{r libraries}
# ── Core packages ─────────────────────────────────────────────────────────────
library(tidyverse)   # data wrangling and ggplot2
library(pwr)         # analytical power functions (pwr.t.test, pwr.2p.test)
library(knitr)       # kable for formatted tables
library(kableExtra)  # extended table styling
library(sandwich)    # cluster-robust variance estimation (vcovCL)
library(lmtest)      # coeftest() with heteroskedasticity/cluster-robust SEs

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold"),
      plot.subtitle    = element_text(color = "grey40"),
      panel.grid.minor = element_blank()
    )
)
```


# Study Design

**Two-stage randomisation:**

**Stage 2 — Individual student-level (chatbot):** All students in the 450 treatment schools are individually randomized 50/50 to chatbot access vs. no chatbot, stratified by school CSE arm. 90 surveyed students per school (30 per grade × 3 grades), yields 40,500 students in the chatbot randomization.


# Key Parameters

```{r parameters}
# Study design 
n_schools_treatment   <- 450    # total treatment schools
n_schools_per_arm     <- 225    # per school-level arm (Regular CSE / CSE+Debiasing)
n_students_per_school <- 90     # surveyed per school (30/grade × 3 grades)

# Chatbot randomization (individual-level 50/50 within each treatment school)
n_chatbot_total   <- n_schools_treatment * n_students_per_school  # 40,500
n_per_chatbot_arm <- n_chatbot_total / 2                           # 20,250
n_per_cell_2x2    <- n_chatbot_total / 4                           # 10,125

# Girls-only subsamples (assume 50% female)
n_girls_per_arm  <- n_per_chatbot_arm / 2   # 10,125
n_girls_per_cell <- n_per_cell_2x2 / 2      # 5,062

# Statistical conventions 
alpha        <- 0.05
power_target <- 0.80
z_alpha      <- qnorm(1 - alpha / 2)   # 1.960
z_beta       <- qnorm(power_target)    # 0.842

# Intra-cluster correlation coefficients (ICC)
# ICC = share of total variance that lies between schools.

icc_pregnancy      <- 0.050   # ENHOGAR national survey 
icc_overconfidence <- 0.065   # 2024 pilot study (N = 800 students)
icc_test_scores    <- 0.200   # Pruebas Nacionales administrative data
icc_depression     <- 0.100   # school-level health literature; pilot study
icc_knowledge      <- 0.100   # within-school component for knowledge/behaviour index

# Outcome-specific baseline parameters 
# Binary outcomes: pregnancy rate
p0_preg_1yr <- 0.056   # annual hazard rate, girls 15–19 (WB WDI; Dupas 2018 SD)
p0_preg_4yr <- 0.168   # four-year cumulative rate

# Continuous outcomes: overconfidence score (0–100 scale)
mean_overconf <- 10.4   # mean from 2024 pilot
sd_overconf   <- 27.7   # SD from 2024 pilot

# Standardized or Bernoulli outcomes (normalized SD = 1 or 0.5)
sd_test_scores <- 1.0   # Pruebas Nacionales, normalized
sd_depression  <- 1.0   # standardized index
sd_knowledge   <- 0.5   # Bernoulli SD at p0 = 0.50
p0_knowledge   <- 0.50  # baseline knowledge/behaviour rate

# Print summary 
tibble(
  Parameter = c(
    "Treatment schools", "Per school-level arm (CSE / CSE+Debiasing)",
    "Surveyed students per school",
    "Total students in chatbot randomisation",
    "Per chatbot arm — Beta_2 (full sample)",
    "Per chatbot arm — girls only (50% female)",
    "Per 2×2 cell — Beta_3 (full sample)",
    "Per 2×2 cell — girls only",
    "Alpha (two-tailed)", "Target power",
    "ICC — pregnancy", "ICC — overconfidence score",
    "ICC — test scores", "ICC — depression/stress",
    "ICC — knowledge/behaviour",
    "Pregnancy base rate — 1 year", "Pregnancy base rate — 4 years",
    "Overconfidence mean (pilot)", "Overconfidence SD (pilot)"
  ),
  Value = c(
    n_schools_treatment, n_schools_per_arm,
    n_students_per_school,
    n_chatbot_total,
    n_per_chatbot_arm, n_girls_per_arm,
    n_per_cell_2x2,    n_girls_per_cell,
    alpha, power_target,
    icc_pregnancy, icc_overconfidence,
    icc_test_scores, icc_depression, icc_knowledge,
    p0_preg_1yr, p0_preg_4yr,
    mean_overconf, sd_overconf
  )
) |>
  kable(caption = "Table 1: Key design and outcome parameters", align = c("l","r")) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) |>
  pack_rows("Study design",            1,  8) |>
  pack_rows("Statistical conventions", 9, 10) |>
  pack_rows("ICC estimates",          11, 15) |>
  pack_rows("Outcome baselines",      16, 19)
```

---

# Analytical Framework: School Fixed Effects and Within-School Variance

```{r within_school_variance}
#  The residual variance driving precision is the *within-school* component:
#   sigma^2_within = sigma^2_total × (1 – ICC)


within_school_sd <- function(sigma_total, icc) sigma_total * sqrt(1 - icc)

tibble(
  Outcome     = c("Pregnancy — 1 year", "Pregnancy — 4 years",
                  "Overconfidence score", "Test scores",
                  "Depression / stress", "Knowledge / behaviour"),
  sigma_total = c(sqrt(p0_preg_1yr * (1 - p0_preg_1yr)),
                  sqrt(p0_preg_4yr * (1 - p0_preg_4yr)),
                  sd_overconf, sd_test_scores, sd_depression, sd_knowledge),
  icc         = c(icc_pregnancy, icc_pregnancy, icc_overconfidence,
                  icc_test_scores, icc_depression, icc_knowledge)
) |>
  mutate(
    sigma_within      = round(within_school_sd(sigma_total, icc), 4),
    variance_kept_pct = round((1 - icc) * 100, 1),
    variance_removed_pct = round(icc * 100, 1)
  ) |>
  rename(
    `σ_total`             = sigma_total,
    `ICC`             = icc,
    `σ_within (with FE)`  = sigma_within,
    `Variance kept (%)`   = variance_kept_pct,
    `Variance removed (%)` = variance_removed_pct
  ) |>
  kable(digits = 4,
        caption = "Table 2: Within-school variance after absorbing school fixed effects") |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

# Analytical MDEs by Outcome

```{r mde_function}
# Returns MDEs and Cohen's d (SE effects) for Estimates 1 and 2.
# For binary outcomes, pass sigma_total = sqrt(p0*(1-p0)).
# For standardized outcomes, pass sigma_total = 1 (SD).

compute_mde <- function(sigma_total, icc,
                        n_arm, n_cell = n_arm / 2,
                        alpha = 0.05, power = 0.80) {
  z_a   <- qnorm(1 - alpha / 2)
  z_b   <- qnorm(power)
  sig_w <- within_school_sd(sigma_total, icc)

  mde1 <- (z_a + z_b) * sig_w * sqrt(2 / n_arm)
  mde2 <- (z_a + z_b) * sig_w * 2 / sqrt(n_cell)

  list(
    sigma_within = sig_w,
    mde_e1       = mde1,           # MDE for Beta2
    mde_e2       = mde2,           # MDE for Beta3
    d_e1         = mde1 / sig_w,   # Cohen's d — Estimate 1 --> In Standard Deviations
    d_e2         = mde2 / sig_w    # Cohen's d — Estimate 2 --> In Standard Deviations
  )
}
```

```{r compute_all_outcomes}
# Apply to all outcomes 

# Pregnancy (binary)
pg1 <- compute_mde(sqrt(p0_preg_1yr * (1 - p0_preg_1yr)), icc_pregnancy,
                   n_per_chatbot_arm, n_per_cell_2x2)
pg4 <- compute_mde(sqrt(p0_preg_4yr * (1 - p0_preg_4yr)), icc_pregnancy,
                   n_per_chatbot_arm, n_per_cell_2x2)

# Overconfidence score (continuous, scale points)
oc  <- compute_mde(sd_overconf,   icc_overconfidence,
                   n_per_chatbot_arm, n_per_cell_2x2)

# Test scores — full sample and girls only
ts_full  <- compute_mde(sd_test_scores, icc_test_scores,
                        n_per_chatbot_arm, n_per_cell_2x2)
ts_girls <- compute_mde(sd_test_scores, icc_test_scores,
                        n_girls_per_arm,  n_girls_per_cell)

# Depression / stress — full sample and girls only
dep_full  <- compute_mde(sd_depression, icc_depression,
                         n_per_chatbot_arm, n_per_cell_2x2)
dep_girls <- compute_mde(sd_depression, icc_depression,
                         n_girls_per_arm,  n_girls_per_cell)

# Knowledge / behaviour index — full sample and girls only
kb_full  <- compute_mde(sd_knowledge, icc_knowledge,
                        n_per_chatbot_arm, n_per_cell_2x2)
kb_girls <- compute_mde(sd_knowledge, icc_knowledge,
                        n_girls_per_arm,  n_girls_per_cell)
```

---

# Summary Table: MDEs for Estimands 1 and 2 {.tabset}

The table below is the primary output of this document. Each row is an outcome; the two columns report the MDE for the **main chatbot effect** ($\beta_2$) and for the **chatbot × debiasing interaction** ($\beta_3$).

```{r summary_mde_table}
# Claude helped me with this table format for a tidy export
outcomes_list <- list(pg1, pg4, oc,
                      ts_full, ts_girls,
                      dep_full, dep_girls,
                      kb_full,  kb_girls)

outcome_names <- c(
  "Pregnancy rate — 1 year",
  "Pregnancy rate — 4 years",
  "Overconfidence score",
  "Test scores",
  "Test scores — girls only",
  "Depression / stress",
  "Depression / stress — girls only",
  "Knowledge / behaviour index",
  "Knowledge / behaviour — girls only"
)

# Natural unit labels for each outcome
units <- c("pp", "pp", "pts", "SD", "SD", "SD", "SD", "pp", "pp")

# ICC used for each outcome (for display)
icc_used <- c(icc_pregnancy, icc_pregnancy, icc_overconfidence,
              icc_test_scores, icc_test_scores,
              icc_depression,  icc_depression,
              icc_knowledge,   icc_knowledge)

# N per arm used for Estimand 1 (halved for girls-only rows)
n_arm_used <- c(n_per_chatbot_arm, n_per_chatbot_arm, n_per_chatbot_arm,
                n_per_chatbot_arm, n_girls_per_arm,
                n_per_chatbot_arm, n_girls_per_arm,
                n_per_chatbot_arm, n_girls_per_arm)

# Helper: format MDE with units and Cohen's d
fmt_mde <- function(r, units_label) {
  sprintf("%.4f %s\n(d = %.4f)", r$mde_e1, units_label, r$d_e1)
}
fmt_mde2 <- function(r, units_label) {
  sprintf("%.4f %s\n(d = %.4f)", r$mde_e2, units_label, r$d_e2)
}

mde_tbl <- tibble(
  Outcome                          = outcome_names,
  `ICC (ρ)`                        = icc_used,
  `N per arm (β₂)`                 = scales::comma(n_arm_used),
  `σ_within`                       = round(sapply(outcomes_list, `[[`, "sigma_within"), 4),
  `MDE — β₂ (Chatbot only)`        = mapply(fmt_mde,  outcomes_list, units),
  `MDE — β₃ (Chatbot × Debiasing)` = mapply(fmt_mde2, outcomes_list, units)
)

mde_tbl |>
  kable(
    caption = paste0(
      "Table 3: Minimum Detectable Effects — Estimates 1 and 2 ",
      "(80% power, α = 0.05, two-tailed)"
    ),
    align = c("l","c","r","r","l","l")
  ) |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = TRUE) |>
  pack_rows("Binary outcomes",       1, 2) |>
  pack_rows("Continuous outcomes",   3, 9) |>
  column_spec(5, width = "18em") |>
  column_spec(6, width = "18em") |>
  footnote(
    general = paste0(
      "MDE(β₂) = (z_β + z_{α/2}) × σ_within × √(2/N_arm). ",
      "MDE(β₃) = (z_β + z_{α/2}) × σ_within × 2/√(n_cell). ",
      "σ_within = σ_total × √(1 – ICC). ",
      "MDE(β₃)/MDE(β₂) = 2 by construction for balanced 2×2. ",
      "Girls-only rows assume 50% female (N halved). ",
      "pp = percentage points; pts = scale points; SD = standard deviation units."
    ),
    general_title = "Notes: "
  )
```



# Open Design Variation: Saturation Design (25%, 50%, 75%)

```{r saturation_analytical}
# Analytical information loss comparison 

# Uniform 50/50: all 450 schools at s = 0.50
info_uniform <- n_schools_treatment * n_students_per_school * 0.50 * 0.50

# Three-level saturation: equal thirds (150 schools each) at 25%, 50%, 75%
sat_levels        <- c(0.25, 0.50, 0.75)
schools_per_level <- n_schools_treatment / length(sat_levels)   # 150 schools each

info_saturation <- sum(
  schools_per_level * n_students_per_school * sat_levels * (1 - sat_levels)
)

info_ratio   <- info_saturation / info_uniform   # relative efficiency (< 1)
se_inflation <- 1 / sqrt(info_ratio)             # SE multiplier (> 1); same for all outcomes

info_loss_pct <- (1 - info_ratio) * 100          # % information lost

# Under saturation, MDE scales up by se_inflation for every outcome

# Per-outcome MDE table

sat_tbl <- tibble(
  Outcome = outcome_names,

  # Columns 1–2: design-level quantities — identical for every outcome
  `Info loss`    = sprintf("−%.1f%%", info_loss_pct),
  `SE inflation` = sprintf("×%.4f",   se_inflation),

  # Columns 3–4: MDE for β₂ (main chatbot effect) under each design
  `MDE β₂ — uniform`    = mapply(function(r, u)
    sprintf("%.4f %s", r$mde_e1, u),
    outcomes_list, units),

  `MDE β₂ — saturation` = mapply(function(r, u)
    sprintf("%.4f %s\n(+%.1f%%)", r$mde_e1 * se_inflation, u,
            (se_inflation - 1) * 100),
    outcomes_list, units),

  # Columns 5–6: MDE for β₃ (chatbot × debiasing interaction) under each design
  `MDE β₃ — uniform`    = mapply(function(r, u)
    sprintf("%.4f %s", r$mde_e2, u),
    outcomes_list, units),

  `MDE β₃ — saturation` = mapply(function(r, u)
    sprintf("%.4f %s\n(+%.1f%%)", r$mde_e2 * se_inflation, u,
            (se_inflation - 1) * 100),
    outcomes_list, units)
)

sat_tbl |>
  kable(
    caption = paste0(
      "Table 5: MDE Comparison — Uniform 50/50 vs. Saturation 25%/50%/75% Design ",
      "(80% power, α = 0.05, two-tailed)"
    ),
    align = c("l", "c", "c", "r", "r", "r", "r")
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) |>
  add_header_above(c(
    " "          = 1,
    "Design cost (constant across outcomes)" = 2,
    "Estimate 1: Main chatbot effect (β₂)"  = 2,
    "Estimate 2: Chatbot × Debiasing (β₃)"  = 2
  )) |>
  pack_rows("Binary outcomes",     1, 2) |>
  pack_rows("Continuous outcomes", 3, 9) |>
  footnote(
    general = paste0(
      "Information loss and SE inflation are design-level quantities — ",
      "identical for all outcomes since the saturation assignment is independent of the outcome. ",
      "MDE(saturation) = MDE(uniform) × SE inflation. ",
      "MDE(β₃) = 2 × MDE(β₂) by construction for a balanced 2×2 design. ",
      "% increase in parentheses = (se_inflation − 1) × 100 = +",
      round((se_inflation - 1) * 100, 1), "% for all cells. ",
      "pp = percentage points; pts = scale points; SD = standard deviation units."
    ),
    general_title = "Notes: "
  )
```

# Power Simulations

```{r powersimu_function}
# R translation of the Stata powersimu program 

powersimu <- function(n_clusters, cluster_size, b_spill, rho) {

  # 1. Create clusters: cluster-level error drives ICC
  u_cluster <- rnorm(n_clusters, mean = 0, sd = sqrt(rho))

  # 2. Assign saturation levels randomly from {0, 0.25, 0.50, 0.75}
  saturation <- sample(c(0, 0.25, 0.50, 0.75), n_clusters, replace = TRUE)

  # 3. Expand to individuals
  id_j  <- rep(seq_len(n_clusters), each = cluster_size)
  u_j   <- rep(u_cluster,           each = cluster_size)
  sat_j <- rep(saturation,          each = cluster_size)

  # 4. Individual-level error: e_ij ~ N(0, sqrt(1 - rho))
  e_ij <- rnorm(n_clusters * cluster_size, mean = 0, sd = sqrt(1 - rho))

  # 5. Assign individual treatment based on cluster saturation
  treated <- rbinom(n_clusters * cluster_size, size = 1, prob = sat_j)

  # 6. Generate outcome
  y <- 0.5 * treated + b_spill * sat_j + u_j + e_ij
  df <- data.frame(id_j = factor(id_j), treated, sat_j, y)

  # 7. Cluster-robust regression
  fit      <- lm(y ~ treated + sat_j, data = df)
  vcov_cl  <- sandwich::vcovCL(fit, cluster = ~ id_j)
  coef_tbl <- lmtest::coeftest(fit, vcov = vcov_cl)

  # P-value for saturation coefficient
  p_val <- coef_tbl["sat_j", "Pr(>|t|)"]

  # Rejection indicator
  as.integer(p_val < 0.05)
}
```


```{r study_sim_grid, cache=TRUE}
# Study-specific simulation: 450 schools × 90 students 

set.seed(28052026)
n_reps_study <- 1000  

b_spill_grid <- c(0.10, 0.20, 0.30, 0.40, 0.50)

# Named vector: rho value (one per study outcome group)
rho_scenarios <- c(
  "Pregnancy (ρ = 0.05)"       = icc_pregnancy,
  "Overconfidence (ρ = 0.065)" = icc_overconfidence,
  "Depression (ρ = 0.10)"      = icc_depression,
  "Test scores (ρ = 0.20)"     = icc_test_scores
)

# Run grid: for each ICC scenario × each b_spill value, run n_reps_study simulations
sim_grid <- map_dfr(names(rho_scenarios), function(scenario) {
  rho_val <- rho_scenarios[[scenario]]

  power_vec <- map_dbl(b_spill_grid, function(bs) {
    res <- lapply(1:n_reps_study, function(x)
      powersimu(
        n_clusters   = n_schools_treatment,    # 450
        cluster_size = n_students_per_school,  # 90
        b_spill      = bs,
        rho          = rho_val
      )
    )
    mean(unlist(res))
  })

  tibble(Scenario = scenario, b_spill = b_spill_grid, power = power_vec)
})
```

## Sim Results

```{r sim_results_table}
# Add 80% power reference row via row_spec highlighting.

sim_wide <- sim_grid |>
  mutate(power_fmt = sprintf("%.1f%%", power * 100)) |>
  select(b_spill, Scenario, power_fmt) |>
  pivot_wider(names_from = Scenario, values_from = power_fmt) |>
  rename(`b_spill (SD)` = b_spill)

sim_wide |>
  kable(
    caption = paste0(
      "Table 6: Simulated Power to Detect the Spillover Effect — ",
      "Saturation Design (25% / 50% / 75%), ",
      n_reps_study, " replications per cell"
    ),
    align = "c"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  add_header_above(c(" " = 1, "Simulated power by ICC scenario" = 4)) |>
  footnote(
    general = paste0(
      "Power = fraction of ", n_reps_study, " simulations rejecting H₀: b_spill = 0 ",
      "at α = 0.05 (two-tailed, cluster-robust SEs via sandwich::vcovCL). ",
      "N = ", scales::comma(n_schools_treatment * n_students_per_school), " students ",
      "across ", n_schools_treatment, " schools. ",
      "Saturation levels drawn uniformly from {0, 0.25, 0.50, 0.75} per school. ",
      "Direct treatment effect fixed at 0.5 SD. ",
      "b_spill is in normalised units where σ_total = 1."
    ),
    general_title = "Notes: "
  )
```